Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-03-28 paper/main.tex
could submit this, but needs claire's eyes

\begin{abstract}
While mutation testing has been a topic of academic interest for
decades, it is only recently that ``real-world'' developers, including
industry leaders such as Google and Meta, have adopted mutation
testing.  In this paper we propose a new approach to the development of mutation
testing tools, and in particular the core challenge of
\emph{generating mutants}.  Current practice tends towards two
limited approaches to mutation generation: mutants are either (1)
generated at the bytecode/IR level, and thus neither human readable
nor adaptable to source-level features of languages or projects, or
(2) generated at the source level by language-specific tools that are
hard to write and maintain, and in fact are often abandoned by both
developers and users.  We propose instead that source-level mutation
generation is a special case of \emph{program transformation} in
general, and that adopting this approach allows for a single tool that
can effectively generate source-level mutants for essentially
\emph{any} programming language. Furthermore, by using \emph{parser
  parser combinators} many of the seeming limitations of an
any-language approach can be overcome, without the need to parse
specific languages.  We compare the universal program transformation
approach to mutation to existing tools, and demonstrate the advantages
of using parser parser combinators to improve on a simple regular-expression
based approach to generation.  
\end{abstract}

@@

Moreover, newer languages, including ones with considerable interest
in the development community, often lack mutation testing tools.  Even
languages with a long history may long lack a working mutation testing
tool; e.g., it was not until 2014 that Haskell had such a tool
\cite{mucheck}.  Furthermore, that tool is no longer
supported\footnote{The last commits or package updates were in 2015, and the
primary author confirms that changes to {\tt haskell-src-ext} since
introduced render it unusable at present.}, and it
appears that no successor tool has appeared in the ensuing
years\footnote{There is FitSpec \cite{FitSpec}, which is maintained,
  but FitSpec does not perform traditional mutation testing or produce
  source code (or bytecode) mutants; it rather
  performs black-box type-aware mutation of function return values.}.
Haskell developers (until now, see below) have no option for
performing true mutation testing.

@@

For the most part we expect users to provide build commands, but do
support automatic compilation and comparison of mutants for Python,
Solidity, Fe, Vyper, Swift, Rust, and Java, including Trivial Compiler
Equivalence \cite{TCE} checks for redundancy of mutants.  These
default handlers can always be over-ridden if they do not work with a
particular build setup (for Python, the approach almost always works,
but in the case of multi-file projects in other languages, it ususally
requires some manual setup).

@@

Table \ref{tab:mutationtools} shows the set of tools compared.  We
attempted to identify well-maintained and commonly used open
source tools for each
language chosen.  The tools cover a number of paradigms; for the most
part we did not choose bytecode based tools, since these often make it
difficult or impossible to mutate a single source file.  PIT, however,
provides this functionality and is perhaps the paradigmatic bytecode
based mutation tool.  Only Dextool, for C++, implements the
\emph{mutant schema} \cite{untch1993mutation} approach often advocated
in the research literature, where a single metaprogram is produced
and, e.g., environment variabes are used to control which particular
version of a program actually executes.  Using a schema has the major
advantage of dramatically reducing the time required to compile
mutants.  We speculate, however, that the fact that the approach is
somewhat complex to implement for some kinds of mutants, and
furthermore fails if even one invalid mutant is generated, has
prevented most developers of non-academic tools from adopting it.

@@

We estimated mutant equivalency when possible by examining a sample of
10 randomly selected unkilled mutants for each file.  For some tools,
human inspection of the individual unkilled mutants was not feasible.
However, we were of course able to estimtate the percent of unkilled
mutants that were unkillable for our two modes for all files, and
these numbers can be put in the perspective of rates reported in the
literature.  Reports of equivalent mutant percentages in the
literature sometimes report the percent of \emph{all} mutants
(including killed mutants, which obviously cannot be equivalent) that
are equivalent.  We follow Schuler and Zeller \cite{EquivMut} in
reporting the percentage of \emph{unkilled} mutants that were found to
be equivalent.  Schuler and Zeller found about 45\% of unkilled
mutants of Java programs to be equivalent.  Other reports of lower
rates, e.g, 9\% \cite{offutt1997automatically} or 7\% \cite{TCE}, report the percent of
\emph{all} mutants, and so translate to much higher rates of
equivalency for unkilled mutants.  We therefore suggest that as a
fraction of unkilled mutants, a rate of 30\% or fewer equivalent
mutants is good, and a rate below 10\% is very good.

@@

Our tools produce somewhat lower mutation scores than other tools, and
higher (though still generally good) equivalency rates than some
tools, but clearly fall into the same broad similarity of behavior
reported for high-quality mutation tools in the literature comparing
Java mutation tools.   The tools with very low or zero equivalency rates, furthermore,
are very conservative in mutation generation, and often simply don't
produce many mutants.

We speculate that our somewhat lower scores and possibly higher
equivalenc rates derive from the simple fact that we generate more
mutants than most tools.  The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, to make mutation testing useful for
projects with extremely high quality test suites, or even using formal
verification, has encouraged a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent, but also
sometimes expose subtle weaknesses in a test suite, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al. \cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single score.

@@

Figure~\ref{fig:haskell} shows a simple Haskell program, the
motivating example in the paper presenting the MuCheck \cite{mucheck}
mutation testing tool for Haskell.  As noted in the introduction, to our
knowledge MuCheck is the first (and perhaps only) mutation tool for
Haskell code.  MuCheck has not been updated since 2015.
To see how our approach fares when mutating code in a language for
which 1) it has no specific rules and 2) the assumption of a simple
C-like syntax does not hold, we applied our implementation to this example.

@@

@@

The approach reported in this paper was briefly presented in part in a
tool paper (citation redacted for blinding purposes) that introduced
the idea of using regular expressions to generate mutants for multiple
languages, and provided an evaluation against PIT for a single toy
Java program.  The present paper provides the full idea and context,
provides empirical data for universal mutant generation on real world
programs over four languages,
and introduces the idea of using parser parser combinators to improve
on regular expression-based generation.

@@

>>>>

\begin{abstract}
While mutation testing has been a topic of academic interest for
decades, it is only recently that ``real-world'' developers, including
industry leaders such as Google and Meta, have adopted mutation
testing.  In this paper we propose a new approach to the development of mutation
testing tools, and in particular the core challenge of
\emph{generating mutants}.  Current practice tends towards two
limited approaches to mutation generation: mutants are either (1)
generated at the bytecode/IR level, and thus neither human readable
nor adaptable to source-level features of languages or projects, or
(2) generated at the source level by language-specific tools that are
hard to write and maintain, and in fact are often abandoned by both
developers and users.  We propose instead that source-level mutation
generation is a special case of \emph{program transformation} in
general, and that adopting this approach allows for a single tool that
can effectively generate source-level mutants for essentially
\emph{any} programming language. Furthermore, by using \emph{parser
  parser combinators} many of the seeming limitations of an
any-language approach can be overcome, without the need to parse
specific languages.  We compare the universal program transformation
approach to mutation to existing tools, and demonstrate the advantages
of using parser parser combinators to improve on a simple regular-expression
based approach to generation.  Finally, we show that our approach
can provide effective mutant generation even for a language for which
it lacks any language-specific operators, and that is not very similar
in syntax to any language it has been applied to previously.
\end{abstract}

@@

Moreover, newer languages, including ones with considerable interest
in the development community, often lack mutation testing tools.  Even
languages with a long history may long lack a working mutation testing
tool; e.g., it was not until 2014 that Haskell had such a tool
\cite{mucheck}.  Furthermore, that tool is no longer
supported\footnote{The last commits or package updates were in 2015, and the
primary author confirms that changes to {\tt haskell-src-ext} since
introduced render it unusable at present.}, and it
appears that no successor tool has appeared in the ensuing
years\footnote{There is FitSpec~\cite{FitSpec}, which is maintained,
  but FitSpec does not perform traditional mutation testing or produce
  source code (or bytecode) mutants; it rather
  performs black-box type-aware mutation of function return values.}.
Haskell developers (until now, see below) have no option for
performing true mutation testing.

@@

For the most part we expect users to provide build commands, but do
support automatic compilation and comparison of mutants for Python,
Solidity, Fe, Vyper, Swift, Rust, and Java, including Trivial Compiler
Equivalence~\cite{TCE} checks for redundancy of mutants.  These
default handlers can always be over-ridden if they do not work with a
particular build setup (for Python, the approach almost always works,
but in the case of multi-file projects in other languages, it ususally
requires some manual setup).

@@

Table \ref{tab:mutationtools} shows the set of tools compared.  We
attempted to identify well-maintained and commonly used open
source tools for each
language chosen.  The tools cover a number of paradigms; for the most
part we did not choose bytecode based tools, since these often make it
difficult or impossible to mutate a single source file.  PIT, however,
provides this functionality and is perhaps the paradigmatic bytecode
based mutation tool.  Only Dextool, for C++, implements the
\emph{mutant schema}~\cite{untch1993mutation} approach often advocated
in the research literature, where a single metaprogram is produced
and, e.g., environment variabes are used to control which particular
version of a program actually executes.  Using a schema has the major
advantage of dramatically reducing the time required to compile
mutants.  We speculate, however, that the fact that the approach is
somewhat complex to implement for some kinds of mutants, and
furthermore fails if even one invalid mutant is generated, has
prevented most developers of non-academic tools from adopting it.

@@

We estimated mutant equivalency when possible by examining a sample of
10 randomly selected unkilled mutants for each file.  For some tools,
human inspection of the individual unkilled mutants was not feasible.
However, we were of course able to estimtate the percent of unkilled
mutants that were unkillable for our two modes for all files, and
these numbers can be put in the perspective of rates reported in the
literature.  Reports of equivalent mutant percentages in the
literature sometimes report the percent of \emph{all} mutants
(including killed mutants, which obviously cannot be equivalent) that
are equivalent.  We follow Schuler and Zeller~\cite{EquivMut} in
reporting the percentage of \emph{unkilled} mutants that were found to
be equivalent.  Schuler and Zeller found about 45\% of unkilled
mutants of Java programs to be equivalent.  Other reports of lower
rates, e.g, 9\%~\cite{offutt1997automatically} or 7\%~\cite{TCE}, report the percent of
\emph{all} mutants, and so translate to much higher rates of
equivalency for unkilled mutants.  We therefore suggest that as a
fraction of unkilled mutants, a rate of 30\% or fewer equivalent
mutants is good, and a rate below 10\% is very good.

@@

Our tools produce somewhat lower mutation scores than other tools, and
higher (though still generally good) equivalency rates than some
tools, but clearly fall into the same broad similarity of behavior
reported for high-quality mutation tools in the literature comparing
Java mutation tools.   The tools with very low or zero equivalency rates, furthermore,
are very conservative in mutation generation, and often simply don't
produce many mutants.  In fact, inspection of the non-comment source
lines as measured by CLOC shown in Tables~\ref{tab:table_java1} and
\ref{tab:table_java2} and Tables~\ref{tab:table_rust1} and
\ref{tab:table_rust2} shows that for many programs, LittleDarwin and
Cargo-Mutants generated \emph{much less than one mutant per actual
  line code}.  Such a conservative approach is likely to produce few
or no actionable (unkilled and non-equivalent) mutants for a program
with a very good, high-coverage test suite.  But such suites are
arguably the ones where mutation testing offers the most promise, by
allowing the detection of very subtle remaining holes in tests for
high-criticality software.

We speculate that our somewhat lower scores and possibly higher
equivalenc rates derive from the simple fact that we generate more
mutants than most tools.  The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, to make mutation testing useful for
projects with extremely high quality test suites, or even using formal
verification, has encouraged a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent, but also
sometimes expose subtle weaknesses in a test suite, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single score.

@@

Figure~\ref{fig:haskell} shows a simple Haskell program, the
motivating example in the paper presenting the MuCheck~\cite{mucheck}
mutation testing tool for Haskell.  As noted in the introduction, to our
knowledge MuCheck is the first (and perhaps only) mutation tool for
Haskell code.  MuCheck has not been updated since 2015.
To see how our approach fares when mutating code in a language for
which 1) it has no specific rules and 2) the assumption of a simple
C-like syntax does not hold, we applied our implementation to this example.

@@

\section{Threats to Validity}

Our study only examined 24 total source files, from 24 different projects, across four different
languages.  None of these files were very large, though they were of
sizes common to many source files in projects across GitHub, and none
were less than about 100 non-comment lines of code.  More importantly,
there is no simple objective way to determine if a mutation testing
tool is ``as good as'' another tool, in an even approximate sense.
Previous studies of the topic have considered that so long as a tool
generates a significant number of non-equivalent unkilled mutants, and
produces a mutation score not extremely differnt than for other tools,
it is a useful mutation testing tool.  Further, there is no
established acceptable tradeoff between generating more mutants and
accepting a higher cost of compilation and execution for mutants.

@@

The approach reported in this paper was briefly presented in part in a
tool paper (citation redacted for blinding purposes) that introduced
the idea of using regular expressions to generate mutants for multiple
languages, and provided an evaluation against PIT for a single toy
Java program.  The present paper provides the full idea and context,
provides empirical data for universal mutant generation on real world
programs over four languages,
and introduces the idea of using parser parser combinators to go
beyond regular expression-based generation.

@@

We conclude that, by using a hierarchical, source-based approach,
either relying on a pure-text regular expression definition of
mutation operators or a richer parser parser combinator transformation
definition, it is possibly to implement effective, easily extensible
mutation testing for essentially \emph{all} commonly used programming
languages, and many non-so-commonly-used programming languages, in
fewer lines of code than any single-language tool we examined.  This
approach fundamentally relies on the insight that program mutation is
just a restricted instance of the general problem of program
transformation, where transformations are highly syntactic (rather
than semantics-aware) and highly localized.

As future work, we plan to give users more control over which mutation
operators are used, with controls common to multiple languages.  We
add that our approach in general seems to be a good fit to modern
large software projects, which are often
multilingual~\cite{Multilingual}: a single mutation tool can be used
for an entire project.
